### 2012-10-02 doc/prelim/document/chapters/introduction/introduction.tex
prelim doc updates

In order to achieve this high fidelity, state-of-the-art computing
\index{facilities} must be leveraged in a way that is both efficient
and considerate of hardware-related issues. As scientific computing
moves towards exascale facilities with machines of $O(1,000,000)$
cores already coming on-line, new algorithms to solve these complex
problems must be developed to leverage this new hardware
\citep{kogge_using_2011}. Issues such as resiliency to node failure,
limited growth of memory available per node, and scaling to large
numbers of cores will be pertinent to robust algorithms aimed at this
new hardware. Considering these issues, this preliminary report
proposes the development of a massively parallel stochastic method for
linear problems and a novel stochastic method to advance solution
techniques for nonlinear problems.

@@

The computational resources required to solve such problems are
tremendous. Recent work in modeling coupled fluid flow and solid
material heat and mass transfer in a reactor subsystem, similar to the
same components of the departure from nucleate boiling example above,
was performed as part of analysis of the Department of Energy's
Consortium for Advanced Simulation of Light Water Reactors (CASL)
modeling and simulation hub using the Drekar multiphysics code
developed at Sandia National Laboratories
\citep{pawlowski_drekar_2012}. Using Drekar, multiphysics simulations
have been performed with fully consistent methods for the solution of
nonlinear systems using meshes of $O(\sn{1}{9})$ elements leveraging
$O(100,000)$ cores on leadership class machines. Neutronics components
to be implemented in CASL for multiphysics analysis, such as the
Denovo radiation transport code developed at Oak Ridge National
Laboratory \citep{evans_denovo:_2010}, compute trillions of unknowns
for full core reactor analysis on $O(\sn{1}{9})$ element meshes and
$O(100,000)$ cores as well. Given the large scale and complexity of
these problems, if we aim to advance multiphysics solution techniques,
then we are motivated to advance the solution of complex and general
nonlinear problems exploiting leadership class levels of parallelism.

\section{Hardware-Based Motivation}
\label{sec:hardware_motivation}
As leadership class machines move towards the exascale, new algorithms
must developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational. Organized work is
already moving forward in this area with the Department of Energy's
Advanced Scientific Computing Research office specifically allocating
funding for the next several years to research resilient solver
technologies for exascale facilities
\citep{u.s._department_of_energy_resilient_2012}. Based on the
language in this call for proposals, we can identify key issues that a
set of robust, massively parallel stochastic solvers could provide a
solution for. As machines begin to operate at hundreds of petaflops
peak performance and beyond, restrictions on power availability will
require incredibly high levels of concurrency. Furthermore, this drop
in power consumption will mean increased pressure on memory as memory
per node is expected to stagnate while cores per node is expected to
increase. As the number of cores increases, their clock speed is
expected to stagnate or even decrease to further reduce power
consumption and manufacturing costs.

@@

\section{Research Outline}
\label{sec:research_outline}
For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\citep{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \citep{saad_iterative_2003,
  kelley_iterative_1995}. In between these two areas are a not widely
known group of stochastic methods for solving sparse linear systems
\citep{forsythe_matrix_1950, hammersley_monte_1964,
  halton_sequential_1962, halton_sequential_1994}. In recent years,
these methods have been further developed for transport problems in
the form of Monte Carlo Synthetic-Acceleration (MCSA)
\citep{evans_residual_2003, evans_monte_2009, evans_monte_2012} but
have yet to be applied to more general sparse linear systems commonly
generated by the computational physics community. Compared to other
methods in this regime, MCSA offers three attractive qualities; (1)
the linear problem operator need not be symmetric or
positive-definite, thereby reducing preconditioning complexity, (2)
the stochastic nature of the solution method provides a natural
solution to the issue of resiliency, and (3) is amenable to
parallelization using modern methods developed by the transport
community \citep{wagner_hybrid_2010}. The development of MCSA as a
general linear solver and the development of a parallel MCSA method
will be new and unique features of this work, providing a framework
with which other issues such as resiliency may be addressed in the
future.

>>>>

In order to achieve this high fidelity, state-of-the-art computing
\index{facilities} must be leveraged in a way that is both efficient
and considerate of hardware-related issues. As scientific computing
moves towards exascale facilities with machines of $O(1,000,000)$
cores already coming on-line, new algorithms to solve these complex
problems must be developed to leverage this new hardware
\citep{kogge_using_2011}. Issues such as resiliency to node failure,
limited growth of memory available per node, and scaling to large
numbers of cores will be pertinent to robust algorithms aimed at this
new hardware. Considering these issues, this preliminary report
proposes the development of a massively parallel Monte Carlo method
for linear problems and a novel Monte Carlo method to advance solution
techniques for nonlinear problems.

@@

The computational resources required to solve such problems are
tremendous. Recent work in modeling coupled fluid flow and solid
material heat and mass transfer in a reactor subsystem, similar to the
same components of the departure from nucleate boiling example above,
was performed as part of analysis of the Department of Energy's
Consortium for Advanced Simulation of Light Water Reactors (CASL)
modeling and simulation hub. CASL used the Drekar multiphysics code
developed at Sandia National Laboratories
\citep{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully consistent methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Denovo radiation transport code developed at Oak Ridge National
Laboratory \citep{evans_denovo:_2010}, compute trillions of unknowns
for full core reactor analysis on $O(\sn{1}{9})$ element meshes and
$O(100,000)$ cores as well. Given the large scale and complexity of
these problems, if we aim to advance multiphysics solution techniques,
then we are motivated to advance the solution of complex and general
nonlinear problems exploiting leadership class levels of parallelism.

\section{Hardware-Based Motivation}
\label{sec:hardware_motivation}
As leadership class machines move towards the exascale, new algorithms
must developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational. Organized work is
already moving forward in this area with the Department of Energy's
Advanced Scientific Computing Research office specifically allocating
funding for the next several years to research resilient solver
technologies for exascale facilities
\citep{u.s._department_of_energy_resilient_2012}. Based on the
language in this call for proposals, we can identify key issues for
which a set of robust, massively parallel Monte Carlo solvers could
provide a solution. As machines begin to operate at hundreds of
petaflops peak performance and beyond, restrictions on power
availability will require incredibly high levels of
concurrency. Furthermore, this drop in power consumption will mean
increased pressure on memory as memory per node is expected to
stagnate while cores per node is expected to increase. As the number
of cores increases, their clock speed is expected to stagnate or even
decrease to further reduce power consumption and manufacturing costs.

@@

\section{Research Outline}
\label{sec:research_outline}
For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\citep{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \citep{saad_iterative_2003,
  kelley_iterative_1995}. In between these two areas are a not widely
known group of Monte Carlo methods for solving sparse linear systems
\citep{forsythe_matrix_1950, hammersley_monte_1964,
  halton_sequential_1962, halton_sequential_1994}. In recent years,
these methods have been further developed for radiation transport
problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
\citep{evans_monte_2009, evans_monte_2012} but have yet to be applied
to more general sparse linear systems commonly generated by the
computational physics community. Compared to other methods in this
regime, MCSA offers three attractive qualities; (1) the linear problem
operator need not be symmetric or positive-definite, thereby reducing
preconditioning complexity, (2) the stochastic nature of the solution
method provides a natural solution to the issue of resiliency, and (3)
is amenable to parallelization using modern methods developed by the
transport community \citep{wagner_hybrid_2010}. The development of
MCSA as a general linear solver and the development of a parallel MCSA
method will be new and unique features of this work, providing a
framework with which other issues such as resiliency may be addressed
in the future.

### 2012-10-02 doc/prelim/document/chapters/research_proposal/research_proposal.tex
prelim doc updates

\section{Proposed Challenge Problem}
\label{sec:challenge_problem}
Our numerical experiments will provide us with a set of metrics giving
best-performance stochastic solver parameters for problems of
different types. As an additional bonus, these experiments will let us
further develop new parallel implementations of the stochastic methods
in order to improve their performance and scalability. With these
advancements in hand, we look forward towards applying them to
large-scale problems in reactor physics. The Consortium for Advanced
Simulation of Light Water Reactors (CASL) is a Department of Energy
Modeling and Simulation Hub focused on predictive multiphysics
modeling of nuclear reactors
\citep{u.s._department_of_energy_casl_2011}. Their efforts are focused
on enabling nuclear reactor analysis and development using today's
supercomputing technology to achieve higher power uprates, higher fuel
burn-up, and predictive accident scenario analysis. As a means of
driving their research and development, CASL has worked with the
nuclear industry to identify challenge problems of critical importance
to achieving these goals.

@@

We culminate our research and development with a challenge problem by
applying these new stochastic methods in the Drekar multiphysics
code. With Drekar, we look forward to providing a new and robust
solution method at the production scale. With these tools and a
leadership class computing platform, we will provide performance
measurements for the stochastic solvers and compare them to
conventional methods currently leveraged within Drekar. With the
potential to improve computational performance in production
environments such as Drekar, these new methods have the potential to
help meet the high fidelity modeling and simulation needs of the
nuclear community.

>>>>

\section{Proposed Challenge Problem}
\label{sec:challenge_problem}
Our numerical experiments will provide us with a set of metrics giving
best-performance Monte Carlo solver parameters for problems of
different types. As an additional bonus, these experiments will let us
further develop new parallel implementations of the Monte Carlo methods
in order to improve their performance and scalability. With these
advancements in hand, we look forward towards applying them to
large-scale problems in reactor physics. The Consortium for Advanced
Simulation of Light Water Reactors (CASL) is a Department of Energy
Modeling and Simulation Hub focused on predictive multiphysics
modeling of nuclear reactors
\citep{u.s._department_of_energy_casl_2011}. Their efforts are focused
on enabling nuclear reactor analysis and development using today's
supercomputing technology to achieve higher power uprates, higher fuel
burn-up, and predictive accident scenario analysis. As a means of
driving their research and development, CASL has worked with the
nuclear industry to identify challenge problems of critical importance
to achieving these goals.

@@

We culminate our research and development with a challenge problem by
applying these new Monte Carlo methods in the Drekar multiphysics
code. With Drekar, we look forward to providing a new and robust
solution method at the production scale. Using these tools and a
leadership class computing platform, we will provide performance
measurements for the Monte Carlo solvers and compare them to
conventional methods currently leveraged within Drekar. Given the
potential to improve computational performance in production
environments such as Drekar, these new methods have the potential to
help meet the high-fidelity modeling and simulation needs of the
nuclear community.
